---
title: "EDA Clustering Tennis"
author: "Jordi Cortés"
format: 
  html:
    toc: true
embed-resources: true
---

```{r}
#| echo: false
#| warning: false
#| message: false

rm(list=ls())

# knitr::opts_knit$set(echo=FALSE)

##-- Load packages -------------------------------------------------------------
library(data.table)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)

##-- Paths ---------------------------------------------------------------------
path_des <- '../Data/disaggregated/'
path_agr <- '../Data/raw_aggregated/'
path_fin <- '../Data/final_data/'
path_fig <- '../Data/Reports/figures/'
path_rep <- '../Data/Reports/'
path_rda <- '../Data/Rdata/'

file_namem <- paste0(path_fin,'d_match.txt')
file_namep <- paste0(path_fin,'d_player.txt')
dm <-  fread(file_namem, header = T, sep=" ") # file matches
dp <-  fread(file_namep, header = T, sep=" ") # file players
```

## Summary

This document shows a preliminary Exploratory Data Analysis (EDA) regarding 
to data from matches of the four Grand Slams in year 2017.

Two main analysis:
- Aggregating by match
- Aggregating by player

## EDA of matches

### Sample size

- Total number of **male** matches: **`r sum(dm$Gender=='M')`**
- Total number of **female** matches: **`r sum(dm$Gender=='W')`**

Table of matches by surface stratified by gender:

```{r}
#| echo: false
t_surfm <- table(dm$surface,dm$Gender)
kable(t_surfm)
```
Percentages:

```{r}
#| echo: false
kable(100*prop.table(t_surfm,2),digits = 1)
```


Table of matches by Slam stratified by gender:

```{r}
#| echo: false
t_slamm <- table(dm$Slam,dm$Gender)
kable(t_slamm)
```
Percentages:

```{r}
#| echo: false
kable(100*prop.table(t_slamm,2),digits = 1)
```

### Performance

Performance by sex.

```{r}
with(dm,tapply(RallyCount,Gender,summary)) # Number of shots. Not Rallies!
with(dm,tapply(Dist,Gender,summary))       # Mean Distance of both Players
with(dm,tapply(Point,Gender,summary))      # Number of points
with(dm,tapply(Speed,Gender,summary))      # Speed serve mean during all the match
with(dm,tapply(PercFirstServe,Gender,summary)) # Percentage of 1st serve for both players
with(dm,tapply(PercSecondServe,Gender,summary))# Percentage of 2nd serve for both players
with(dm,tapply(PercDoubleFault,Gender,summary))# Percentage of double faults for both players
with(dm,tapply(total_minutes,Gender,summary))   # Total minutes of the match
with(dm,tapply(played_minutes,Gender,summary))  # Played minutes of the match (20% of the total minutes)
with(dm,tapply(rally_min,Gender,summary))       # shots per minute
with(dm,tapply(Dist_min,Gender,summary))        # Distance for any player per real (played) minute 
```



## EDA of players

### Sample size

- Total number of **male** players: **`r sum(dp$Gender=='M')`**
- Total number of **female** players: **`r sum(dp$Gender=='W')`**

Table of number of matches by player stratified by gender:

```{r}
#| echo: false
t_num_m <- table(dp$n,dp$Gender)
kable(t_num_m)
```
Percentages:

```{r}
#| echo: false
kable(100*prop.table(t_num_m,2),digits = 1)
```

### Performance

Performance by sex.

```{r}
# with(dp,tapply(RallyCount,Gender,summary))   # Number of shots in all matches. Not Rallies!
# with(dp,tapply(PDist,Gender,summary))        # Distance Player in all matches
# with(dp,tapply(Point,Gender,summary))        # Number of points in all matches
with(dp,tapply(Speed,Gender,summary))          # Speed serve mean during all the match
with(dp,tapply(PercFirstServe,Gender,summary)) # Percentage of 1st serve player
with(dp,tapply(PercSecondServe,Gender,summary))# Percentage of 2nd serve player
with(dp,tapply(PercDoubleFault,Gender,summary))# Percentage of double faults serve player
with(dp,tapply(rally_min,Gender,summary))      # shots per minute
with(dp,tapply(Dist_min,Gender,summary))       # player distance (meters) per minute
with(dp,tapply(rally_match,Gender,summary))    # shots per match
with(dp,tapply(Dist_match,Gender,summary))     # player distance (meters) per match
```

## Appendix

### Aggregated data by match (only male)

```{r}
#| echo: false
kable(dm[Gender=='M'],digits = 1) %>% kable_styling()
```


### Aggregated data by player (only male)

**n**: number of matches

```{r}
#| echo: false
dp_m <- dp[Gender=='M']
kable(dp_m[order(dp_m$n, decreasing = TRUE),],digits = 1) %>% kable_styling()
```

