---
title: "Grand Slams 2017. Clustering Analysis"
author: "Jordi Cortés"
format: 
  html:
    toc: true
embed-resources: true
---

```{r}
#| echo: false
#| warning: false
#| message: false

rm(list=ls())

# knitr::opts_knit$set(echo=FALSE)

##-- Load packages -------------------------------------------------------------
library(data.table)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gtsummary)
library(compareGroups)
library(archetypes)
library(adamethods)

##-- Paths ---------------------------------------------------------------------
path_des <- '../Data/disaggregated/'
path_agr <- '../Data/raw_aggregated/'
path_fin <- '../Data/final_data/'
path_fig <- 'figures/'
path_rep <- '../Data/Reports/'
path_rda <- '../Data/Rdata/'

##-- Functions -----------------------------------------------------------------
source('adaord/code/auxiliary/LFunctions_to_calculate_real_archetypes_with_swap2.R')
source('adaord/code/auxiliary/LstepArchetypesMod.R')
source('adaord/code/auxiliary/LstepArchetypoids.R')

file_namem <- paste0(path_fin,'d_match.txt')
file_namep <- paste0(path_fin,'d_player.txt')
dm <-  fread(file_namem, header = T, sep=" ") # file matches
dp <-  fread(file_namep, header = T, sep=" ") # file players
```

## Executive summary

This document shows the results of a clustering analysis applied to matches of the **four Grand Slam men's singles in year 2017**.

There are two main clustering analysis:

- Classifying **matches** according to their characteristics of workload and intensity 
- Classifying **players** according to their characteristics of workload and intensity in the matches that they played

## Descriptive of the sample

### Matches

The total collected number of matches was **`r sum(dm$Gender=='M')`** matches.

```{r}
#| echo: false
dm2 <- dm |> filter(Gender=='M') |> select(surface, Slam)
dm2 |> tbl_summary()
```

### Players

The total number of collected male players participated in Grand Slams during 2017 was **`r sum(dp$Gender=='M')`**.

### Variables

The collected variables for each match were: 

*`r paste(names(dm), collapse=', ')`*

The collected variables for each player were: 

*`r paste(names(dp), collapse=', ')`*


```{r}
#| echo: false

# SELECT relevant variables

##-- Matches ----------------------------------------------------------------
##-- Variables for performing the clustering in matches
var_clust_m <- c('played_minutes','Point','RallyCount','Dist', # Load variables
               'rally_min','Dist_min')                       # Intensity  
##-- Variables to compare between clusterings in matches
var_int_m   <- c(var_clust_m,'Speed','surface',
               'PercFirstServe','PercSecondServe','PercDoubleFault') # Intensity 
# scaled clustering --> Matches
dm_m <- dm %>% filter(Gender=='M') %>% select(all_of(var_clust_m)) %>% scale()# Men
# Not scaled descriptive --> Matches
dm_m0 <- dm %>% filter(Gender=='M') %>% select(all_of(var_int_m)) # Men

##-- Players ----------------------------------------------------------------
##-- Variables for performing the clustering in players
# var_clust_p <- c('played_minutes','Point','RallyCount','Dist','rally_match','Dist_match', # Load variables
#                 'rally_min','Dist_min')# Intensity  
var_clust_p <- c('played_minutes_match','Point_match','rally_match','Dist_match', # Load variables
                 'rally_min','Dist_min')# Intensity

var_int_p   <- c(var_clust_p,'Speed',
               'PercFirstServe','PercSecondServe','PercDoubleFault','Player') # Intensity 
# scaled clustering --> Players
dp_m <- dp %>% filter(Gender=='M') %>% select(all_of(var_clust_p)) %>% scale()# Men
# Not scaled descriptive --> Players
dp_m0 <- dp %>% filter(Gender=='M') %>% select(all_of(var_int_p)) # Men

```


## Clustering of matches

### Methods

K-means algorithm was used to perform this analysis. We use the following variables build the clusters: `r paste(colnames(dm_m),collapse=', ')`


### Number of clusters

We use the elbow rule to determine the number of clusters. There are **4** clusters.

```{r echo=FALSE}

#| echo: false

################################################################################
# Clustering per match --> Men
################################################################################
set.seed(12345)

# Elbow men --> k=4
k.max <- 10
wss <- sapply(1:k.max,function(k){kmeans(dm_m, k, nstart=50, iter.max = 15 )$tot.withinss})
plot(1:k.max, wss,type="b", pch = 19, frame = FALSE,xlab="Number of clusters K", ylab="Total within-clusters sum of squares")
abline(v=4,lty=3)
```


### Cluster description

Next table describe each of the clusters according to their characteristics.

```{r}
#| echo: false

# Number of clusters=4
km_m <- kmeans(dm_m, centers =  4, nstart=50, iter.max = 15)
dm_m0$cluster <- km_m$cluster
res <- compareGroups(cluster ~ ., data=dm_m0)
createTable(res)
```


### Interpretation

Long matches are measured with four variables: 

- __*played_minutes*__. Number of effectively played minutes during the match (20% of the total time).    
- __*Point*__. Number of points during the match. 
- __*RallyCount*__. Number of shots (not rallies!) during the match.  
- __*Dist*__. Average distance (meters) covered by the two players of the match.

On the other hand, intensity is measured with two variables:

- __*rally_min*__. Number of shots per minute.
- __*Dist_min*__. Average distance (meters) covered by the two players of the match per minute.

There are four types of matches:

- **Cluster 1** $\rightarrow$ Short matches with low intensity. There is a high proportion of matches in **Grass** surface (41%, which is a 57% more than in the original sample). On the other hand, there is a low presence of **Hard** surface (44%, which is a 25% less than in the original sample)

- **Cluster 2** $\rightarrow$ Short matches with high intensity. There is a high proportion of matches in **Clay** surface (22%, which is a 57% more than in the original sample). 

- **Cluster 3** $\rightarrow$ Long matches with high intensity. There is a high proportion of matches in **Hard** surface (80%, which is a 36% more than in the original sample). On the other hand, there is a low presence of **Grass** surface (8%, which is a 69% less than in the original sample)


- **Cluster 4** $\rightarrow$ Long matches with low intensity. There is a low presence of **Clay** surface (8%, which is a 43% less than in the original sample)


There are no relevant differences regarding the percentages of first and second serves and double faults.


## Clustering of players

### Methods

We performed an archetypoidal analysis to find the most representative players from each cluster. We used the following variables to define the archetypoids: `r paste0(colnames(dp_m),collapse=', ')`

### Number of clusters

```{r}
#| echo: false
#| message: false
#| warning: false

##-- Archetypes
set.seed(12345)
norep_test <- 5
norep      <- 10

## Previously run
# AA  <- stepLArchetypoids3(data=dp_m, k=1:10,   norep = norep_test)
# AA3 <- stepLArchetypoids3(data=dp_m, k=3, norep = norep)
# AA4 <- stepLArchetypoids3(data=dp_m, k=4, norep = norep)
# save(AA,AA3,AA4, file = '../Data/Rdata/Archetypoids.Rdata')
load('../Data/Rdata/Archetypoids.Rdata')
```

The screeplot say that there are 3 o 4 player's archetipoids.

```{r}
#| echo: false

# Elbow rule
screeplot(AA)
abline(v=4,lty=2)
```


### Cluster description

We decided to take 4 clusters according to theirs similarities to the archetipoids. Next table describes each of the clusters according to their characteristics.

```{r}
#| echo: false
#| warning: false


##-- Number archetypoids -------------------------------------------------------
num_arq <- 4 
n_arquetipoides     <- AA[[num_arq]][[1]]         # n_arquetipoides contains the individual numbers which are archetypoids
player_archetypoids <- dp$Player[n_arquetipoides] # arquetipoid players


##-- Alphas --------------------------------------------------------------------
ua3r <- dp_m[n_arquetipoides,]
huge <- 200
k    <- 4
n    <- ncol(t(dp_m))
x_gvv<- rbind(t(dp_m), rep(huge, n))
zs   <- x_gvv [,n_arquetipoides]
ae   <- matrix(0, nrow = k, ncol = n)
for (j in 1 : n){
  ae[, j] = coef(nnls(zs, x_gvv[,j]))
}
ae <- t(ae)

##-- Assign to clusters --------------------------------------------------------
dp_m0$cluster <- apply(ae,1,which.max)
res           <- compareGroups(cluster ~ ., data=dp_m0)
createTable(res)
```


### Interpretation

```{r}
#| echo: false
sel_1 <- which(dp_m0$cluster==1)
sel_2 <- which(dp_m0$cluster==2)
sel_3 <- which(dp_m0$cluster==3)
sel_4 <- which(dp_m0$cluster==4)
```

```{r}
#| echo: false

# Most representatives
pp_pos1 <- which.max(ae[,1]) # Prototipus cluster 1
pp_pos2 <- which.max(ae[,2]) # Prototipus cluster 2
pp_pos3 <- which.max(ae[,3]) # Prototipus cluster 3
pp_pos4 <- which.max(ae[,4]) # Prototipus cluster 4
```

The duration of the match is measured with variables:

- _**played_minutes_match**_: minuts per partit
- _**Point_match**_: punts per partit
- _**rally_match**_: cops per partit
- _**Dist_match**_: distancia per partit

The intensity of the match is measured with variables:
- _**rally_min**_: cops per minut
- _**Dist_min**_: distancia per minut



According to this, we can define the following clusters:

- **Cluster 1** $\rightarrow$ Moderate-long matches with high intensity. The player most representative is **`r dp_m0$Player[pp_pos1]`**

- **Cluster 2** $\rightarrow$ Moderate-short with low intensity. The player most representative is **`r dp_m0$Player[pp_pos2]`**

- **Cluster 3** $\rightarrow$ Short matches with moderate-high intensity. The player most representative is **`r dp_m0$Player[pp_pos3]`**

- **Cluster 4** $\rightarrow$ Long matches with moderate-low intensity. The player most representative is **`r dp_m0$Player[pp_pos4]`**

There are no differences regarding the percentages of first/second serves between clusters.

### Composition of each player

Each player that participated in some Grand Slam in 2017 is a linear combination of the four
next four archetipoids: 

- Black: _**Darian King**_
- Red:   _**Sam Groth**_
- Green: _**Luca Vanni**_
- Blue:  _**Janko Tipsarevic**_

**Cluster 1**

Players more similars to **`r dp_m0$Player[pp_pos1]`**

```{r}
#| echo: false

cex_lab <- 0.5

stars(ae[sel_1,],draw.segments=T,labels=dp_m0$Player[sel_1],cex=cex_lab,scale=FALSE)
```


**Cluster 2**

Players more similars to **`r dp_m0$Player[pp_pos2]`**

```{r}
#| echo: false
stars(ae[sel_2,],draw.segments=T,labels=dp_m0$Player[sel_2],cex=cex_lab,scale=FALSE)
```

**Cluster 3**

Players more similars to **`r dp_m0$Player[pp_pos3]`**

```{r}
#| echo: false
stars(ae[sel_3,],draw.segments=T,labels=dp_m0$Player[sel_3],cex=cex_lab,scale=FALSE)
```

**Cluster 4**

Players more similars to **`r dp_m0$Player[pp_pos4]`**

```{r}
#| echo: false
stars(ae[sel_4,],draw.segments=T,labels=dp_m0$Player[sel_4],cex=cex_lab,scale=FALSE)
```

## Appendix I: Clustering of matches

```{r}
#| echo: false

dm_list <- dm %>% filter(Gender=='M') %>% select(all_of(c('match_id','player1','player2','Slam',var_clust_m))) 
dm_list$cluster <- km_m$cluster
```

### Cluster 1

```{r}
#| echo: false
dm_list_temp <- dm_list %>% filter(cluster==1)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 2

```{r}
#| echo: false
dm_list_temp <- dm_list %>% filter(cluster==2)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 3

```{r}
#| echo: false
dm_list_temp <- dm_list %>% filter(cluster==3)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 4

```{r}
#| echo: false
dm_list_temp <- dm_list %>% filter(cluster==4)
kable(dm_list_temp, digits=1) %>% kable_styling()
```


## Appendix II: Clustering of players


```{r}
#| echo: false

dp_list <- dp %>% filter(Gender=='M') %>% select(all_of(c('Player',var_clust_p))) 
dp_list$cluster <- dp_m0$cluster
```

### Cluster 1

```{r}
#| echo: false
dm_list_temp <- dp_list %>% filter(cluster==1)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 2

```{r}
#| echo: false
dm_list_temp <- dp_list %>% filter(cluster==2)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 3

```{r}
#| echo: false
dm_list_temp <- dp_list %>% filter(cluster==3)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 4

```{r}
#| echo: false
dm_list_temp <- dp_list %>% filter(cluster==4)
kable(dm_list_temp, digits=1) %>% kable_styling()
```


