---
title: "Grand Slams 2017. Clustering Analysis"
author: "Jordi Cortés"
format: 
  html:
    toc: true
embed-resources: true
---

```{r}
#| echo: false
#| warning: false
#| message: false

rm(list=ls())

# knitr::opts_knit$set(echo=FALSE)

##-- Load packages -------------------------------------------------------------
library(data.table)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(gtsummary)
library(compareGroups)
library(archetypes)

##-- Paths ---------------------------------------------------------------------
path_des <- '../Data/disaggregated/'
path_agr <- '../Data/raw_aggregated/'
path_fin <- '../Data/final_data/'
path_fig <- 'figures/'
path_rep <- '../Data/Reports/'
path_rda <- '../Data/Rdata/'

file_namem <- paste0(path_fin,'d_match.txt')
file_namep <- paste0(path_fin,'d_player.txt')
dm <-  fread(file_namem, header = T, sep=" ") # file matches
dp <-  fread(file_namep, header = T, sep=" ") # file players
```

## Executive summary

This document shows the results of a clustering analysis applied to matches of the **four Grand Slam men's singles in year 2017**.

There are two main clustering analysis:

- Classifying **matches** according to their characteristics of workload and intensity 
- Classifying **players** according to their characteristics of workload and intensity in the matches that they played

## Descriptive of the sample

### Matches

The total collected number of matches was **`r sum(dm$Gender=='M')`** matches.

```{r}
#| echo: false
dm2 <- dm |> filter(Gender=='M') |> select(surface, Slam)
dm2 |> tbl_summary()
```

### Players

The total number of collected male players participated in Grand Slams during 2017 was **`r sum(dp$Gender=='M')`**.

### Variables

The collected variables for each match were: 

*`r paste(names(dm), collapse=', ')`*

The collected variables for each player were: 

*`r paste(names(dp), collapse=', ')`*


```{r}
#| echo: false

# SELECT relevant variables

##-- Matches ----------------------------------------------------------------
##-- Variables for performing the clustering in matches
var_clust_m <- c('played_minutes','Point','RallyCount','Dist', # Load variables
               'rally_min','Dist_min')                       # Intensity  
##-- Variables to compare between clusterings in matches
var_int_m   <- c(var_clust_m,'Speed','surface',
               'PercFirstServe','PercSecondServe','PercDoubleFault') # Intensity 
# scaled clustering --> Matches
dm_m <- dm %>% filter(Gender=='M') %>% select(all_of(var_clust_m)) %>% scale()# Men
# Not scaled descriptive --> Matches
dm_m0 <- dm %>% filter(Gender=='M') %>% select(all_of(var_int_m)) # Men

##-- Players ----------------------------------------------------------------
##-- Variables for performing the clustering in matches
var_clust_p <- c('played_minutes','Point','RallyCount','Dist','rally_match','Dist_match', # Load variables
                'rally_min','Dist_min')# Intensity  
# var_clust_p <- c('played_minutes_match','Point_match','rally_match','Dist_match', # Load variables
#                'rally_min','Dist_min')# Intensity  

var_int_p   <- c(var_clust_p,'Speed',
               'PercFirstServe','PercSecondServe','PercDoubleFault','Player') # Intensity 
# scaled clustering --> Players
dp_m <- dp %>% filter(Gender=='M') %>% select(all_of(var_clust_p)) %>% scale()# Men
# Not scaled descriptive --> Players
dp_m0 <- dp %>% filter(Gender=='M') %>% select(all_of(var_int_p)) # Men

```


## Clustering of matches

### Methods

K-means algorithm was used to perform this analysis.

### Number of clusters

We use the elbow rule to determine the number of clusters. There are **4** clusters.

```{r echo=FALSE}

#| echo: false

################################################################################
# Clustering per match --> Men
################################################################################
set.seed(12345)

# Elbow men --> k=4
k.max <- 10
wss <- sapply(1:k.max,function(k){kmeans(dm_m, k, nstart=50, iter.max = 15 )$tot.withinss})
plot(1:k.max, wss,type="b", pch = 19, frame = FALSE,xlab="Number of clusters K", ylab="Total within-clusters sum of squares")
abline(v=4,lty=3)
```


### Cluster description

Next table describe each of the clusters according to their characteristics.

```{r}
#| echo: false

# Number of clusters=4
km_m <- kmeans(dm_m, centers =  4, nstart=50, iter.max = 15)
dm_m0$cluster <- km_m$cluster
res <- compareGroups(cluster ~ ., data=dm_m0)
createTable(res)
```


### Interpretation

Long matches are measured with four variables: 

- __*played_minutes*__. Number of effectively played minutes during the match (20% of the total time).    
- __*Point*__. Number of points during the match. 
- __*RallyCount*__. Number of shots (not rallies!) during the match.  
- __*Dist*__. Average distance (meters) covered by the two players of the match.

On the other hand, intensity is measured with two variables:

- __*rally_min*__. Number of shots per minute.
- __*Dist_min*__. Average distance (meters) covered by the two players of the match per minute.

There are four types of matches:

- **Cluster 1** $\rightarrow$ Short matches with low intensity. There is a high proportion of matches in **Grass** surface (41%, which is a 57% more than in the original sample). On the other hand, there is a low presence of **Hard** surface (44%, which is a 25% less than in the original sample)

- **Cluster 2** $\rightarrow$ Short matches with high intensity. There is a high proportion of matches in **Clay** surface (22%, which is a 57% more than in the original sample). 

- **Cluster 3** $\rightarrow$ Long matches with high intensity. There is a high proportion of matches in **Hard** surface (80%, which is a 36% more than in the original sample). On the other hand, there is a low presence of **Grass** surface (8%, which is a 69% less than in the original sample)


- **Cluster 4** $\rightarrow$ Long matches with low intensity. There is a low presence of **Clay** surface (8%, which is a 43% less than in the original sample)


There are no relevant differences regarding the percentages of first and second serves and double faults.


## Clustering of players

### Methods

We performed an archetypal analysis to find the most representative players from each cluster analysis.

### Number of clusters

The screeplot say that there is 3 o 4 clusters.

```{r}
#| echo: false
#| message: false
#| warning: false

##-- Archetypes
set.seed(12345)
aa <- stepArchetypes(dp_m, k=1:10, nrep=5,verbose = FALSE)
screeplot(aa)
```


### Cluster description

We decided to take 4 clusters. Next table describe each of the clusters according to their characteristics.

```{r}
#| echo: false
#| warning: false


##-- Four clusters -------------------------------------------------------
aa_3 <- bestModel(aa[[3]]) 
aa_4 <- bestModel(aa[[4]]) 

##-- Assign to clusters --> Number of clusters=4
dp_m0$cluster <- apply(coef(aa_4),1,which.max)
res <- compareGroups(cluster ~ ., data=dp_m0)
createTable(res)
```


### Characteristics 

The standarized characteristics of each cluster are:

```{r}
#| echo: false

aa_4$archetypes
```

### Interpretation

```{r}
#| echo: false

# Most representatives
pp_pos1 <- which.max(coef(aa_4)[,1]) # Prototipus cluster 1
pp_pos2 <- which.max(coef(aa_4)[,2]) # Prototipus cluster 2
pp_pos3 <- which.max(coef(aa_4)[,3]) # Prototipus cluster 3
pp_pos4 <- which.max(coef(aa_4)[,4]) # Prototipus cluster 4
```



- **Cluster 1** $\rightarrow$ Not to much matches played with high intensity. The player most representative is **`r dp_m0$Player[pp_pos1]`**

- **Cluster 2** $\rightarrow$ Not to much matches played with moderate intensity. The player most representative is **`r dp_m0$Player[pp_pos2]`**

- **Cluster 3** $\rightarrow$ A lot of  matches played with moderate intensity. The player most representative is **`r dp_m0$Player[pp_pos3]`**

- **Cluster 4** $\rightarrow$ Few matches played with low intensity. The player most representative is **`r dp_m0$Player[pp_pos4]`**


## Appendix I: Clustering of matches

```{r}
#| echo: false

dm_list <- dm %>% filter(Gender=='M') %>% select(all_of(c('match_id','player1','player2','Slam',var_clust_m))) 
dm_list$cluster <- km_m$cluster
```

### Cluster 1

```{r}
#| echo: false
dm_list_temp <- dm_list %>% filter(cluster==1)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 2

```{r}
#| echo: false
dm_list_temp <- dm_list %>% filter(cluster==2)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 3

```{r}
#| echo: false
dm_list_temp <- dm_list %>% filter(cluster==3)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 4

```{r}
#| echo: false
dm_list_temp <- dm_list %>% filter(cluster==4)
kable(dm_list_temp, digits=1) %>% kable_styling()
```


## Appendix II: Clustering of players


```{r}
#| echo: false

dp_list <- dp %>% filter(Gender=='M') %>% select(all_of(c('Player',var_clust_p))) 
dp_list$cluster <- dp_m0$cluster
```

### Cluster 1

```{r}
#| echo: false
dm_list_temp <- dp_list %>% filter(cluster==1)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 2

```{r}
#| echo: false
dm_list_temp <- dp_list %>% filter(cluster==2)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 3

```{r}
#| echo: false
dm_list_temp <- dp_list %>% filter(cluster==3)
kable(dm_list_temp, digits=1) %>% kable_styling()
```

### Cluster 4

```{r}
#| echo: false
dm_list_temp <- dp_list %>% filter(cluster==4)
kable(dm_list_temp, digits=1) %>% kable_styling()
```


